\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-06-29 paper.qmd
Edit page 8

The solution is to use the tour to examine how the NLDR is warping the space. This approach follows what @wickham2015 describes as *model-in-the-data-space*. The fitted model should be overlaid on the data, to examine the fit relative the spread of the observations. While this is straightforward, and commonly done when data is 2D, it is also possible in \pD{}, for many models, when a tour is used.

@@

XXX Model in 2D, on the UMAP layout, and three views of the fit in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape, but it has breaks which means that it does not adequately capture the surface. (The **langevitour** software is used to view the data with a tour, and the full video is available at (<https://youtu.be/G1m4q9k--v4>). NEED TO CHANGE FIRST PLOT TO BE 2D WIREFRAME.
:::

@@

where $n$ is XXX, $b$ is XXX, $n_h$ is XXX, $p$ is XXX, $\mathbfit{x}_{hij}$ is XXX. MAKE THIS A NUMBERED EQUATION, SO WE CAN REFER TO IT LATER

@@

In the 2D layouts generated by various NLDR techniques, as shown in @fig-NLDR-variety-gau, five well-separated clusters are shown. In tSNE (see @fig-NLDR-variety-gau (a)), these clusters appear closely. UMAP arranges all clusters in a parallel manner, with three aligned in one line and the other two in a separate line (see @fig-NLDR-variety-gau (b)). In contrast, PHATE shows two closely positioned clusters and three more distant ones (see @fig-NLDR-variety-gau (c)). In TriMAP, two clusters are close, though not as tightly as PHATE, while the other three are well-separated (see @fig-NLDR-variety-gau (d)). Finally, PaCMAP shows one central cluster and the remaining four spread out in different directions (see @fig-NLDR-variety-gau (e)).

>>>>

The solution is to use the tour to examine how the NLDR is warping the space. This approach follows what @wickham2015 describes as *model-in-the-data-space*. The fitted model should be overlaid on the data, to examine the fit relative the spread of the observations. While this is straightforward, and commonly done when data is \gD{}, it is also possible in \pD{}, for many models, when a tour is used.

@@

XXX Model in \gD{}, on the UMAP layout, and three views of the fit in projections from $7\text{-}D$, for the S-curve data. The model closely fits the shape, but it has breaks which means that it does not adequately capture the surface. (The **langevitour** software is used to view the data with a tour, and the full video is available at (<https://youtu.be/G1m4q9k--v4>). NEED TO CHANGE FIRST PLOT TO BE 2D WIREFRAME.
:::

@@

where $n$ is the number of observations, $b$ is the number of bins, $n_h$ is the number of observations in $h^{th}$ bin, $p$ is the number of variables, $\mathbfit{x}_{hij}$ is the $j^{th}$ dimensional data of $i^{th}$ observation in $h^{th}$ hexagon.

@@

In the \gD{} layouts generated by various NLDR techniques, as shown in @fig-NLDR-variety-gau, five well-separated clusters are shown. In tSNE (see @fig-NLDR-variety-gau (a)), these clusters appear closely. UMAP arranges all clusters in a parallel manner, with three aligned in one line and the other two in a separate line (see @fig-NLDR-variety-gau (b)). In contrast, PHATE shows two closely positioned clusters and three more distant ones (see @fig-NLDR-variety-gau (c)). In TriMAP, two clusters are close, though not as tightly as PHATE, while the other three are well-separated (see @fig-NLDR-variety-gau (d)). Finally, PaCMAP shows one central cluster and the remaining four spread out in different directions (see @fig-NLDR-variety-gau (e)).

### 2024-06-29 paper.tex
Page 8 numbered equation
Edit page 8

The solution is to use the tour to examine how the NLDR is warping the
space. This approach follows what \citet{wickham2015} describes as
\emph{model-in-the-data-space}. The fitted model should be overlaid on
the data, to examine the fit relative the spread of the observations.
While this is straightforward, and commonly done when data is 2D, it is
also possible in \pD{}, for many models, when a tour is used.

@@

\caption{\label{fig-scurve-sc}XXX Model in 2D, on the UMAP layout, and
three views of the fit in projections from \(7\text{-}D\), for the
S-curve data. The model closely fits the shape, but it has breaks which
means that it does not adequately capture the surface. (The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at (\url{https://youtu.be/G1m4q9k--v4}).
NEED TO CHANGE FIRST PLOT TO BE 2D WIREFRAME.}

@@

\[\frac{1}{n}\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2\]

where \(n\) is XXX, \(b\) is XXX, \(n_h\) is XXX, \(p\) is XXX,
\(\mathbfit{x}_{hij}\) is XXX. MAKE THIS A NUMBERED EQUATION, SO WE CAN
REFER TO IT LATER

@@

In the 2D layouts generated by various NLDR techniques, as shown in
Figure~\ref{fig-NLDR-variety-gau}, five well-separated clusters are
shown. In tSNE (see Figure~\ref{fig-NLDR-variety-gau} (a)), these
clusters appear closely. UMAP arranges all clusters in a parallel
manner, with three aligned in one line and the other two in a separate
line (see Figure~\ref{fig-NLDR-variety-gau} (b)). In contrast, PHATE
shows two closely positioned clusters and three more distant ones (see
Figure~\ref{fig-NLDR-variety-gau} (c)). In TriMAP, two clusters are
close, though not as tightly as PHATE, while the other three are
well-separated (see Figure~\ref{fig-NLDR-variety-gau} (d)). Finally,
PaCMAP shows one central cluster and the remaining four spread out in
different directions (see Figure~\ref{fig-NLDR-variety-gau} (e)).

>>>>

The solution is to use the tour to examine how the NLDR is warping the
space. This approach follows what \citet{wickham2015} describes as
\emph{model-in-the-data-space}. The fitted model should be overlaid on
the data, to examine the fit relative the spread of the observations.
While this is straightforward, and commonly done when data is \gD{}, it
is also possible in \pD{}, for many models, when a tour is used.

@@

\caption{\label{fig-scurve-sc}XXX Model in \gD{}, on the UMAP layout,
and three views of the fit in projections from \(7\text{-}D\), for the
S-curve data. The model closely fits the shape, but it has breaks which
means that it does not adequately capture the surface. (The
\textbf{langevitour} software is used to view the data with a tour, and
the full video is available at (\url{https://youtu.be/G1m4q9k--v4}).
NEED TO CHANGE FIRST PLOT TO BE 2D WIREFRAME.}

@@

\begin{equation}\phantomsection\label{eq-equation1}{\frac{1}{n}\sum_{h = 1}^{b}\sum_{i = 1}^{n_h}\sum_{j = 1}^{p} (\mathbfit{x}_{hij} - C^{(p)}_{hj})^2}\end{equation}

where \(n\) is the number of observations, \(b\) is the number of bins,
\(n_h\) is the number of observations in \(h^{th}\) bin, \(p\) is the
number of variables, \(\mathbfit{x}_{hij}\) is the \(j^{th}\)
dimensional data of \(i^{th}\) observation in \(h^{th}\) hexagon.

@@

In the \gD{} layouts generated by various NLDR techniques, as shown in
Figure~\ref{fig-NLDR-variety-gau}, five well-separated clusters are
shown. In tSNE (see Figure~\ref{fig-NLDR-variety-gau} (a)), these
clusters appear closely. UMAP arranges all clusters in a parallel
manner, with three aligned in one line and the other two in a separate
line (see Figure~\ref{fig-NLDR-variety-gau} (b)). In contrast, PHATE
shows two closely positioned clusters and three more distant ones (see
Figure~\ref{fig-NLDR-variety-gau} (c)). In TriMAP, two clusters are
close, though not as tightly as PHATE, while the other three are
well-separated (see Figure~\ref{fig-NLDR-variety-gau} (d)). Finally,
PaCMAP shows one central cluster and the remaining four spread out in
different directions (see Figure~\ref{fig-NLDR-variety-gau} (e)).
